---
title: "Pairs Trading using the Kalman Filter"
author: "Lucas S. Macoris"
date: "14/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## **About this document**

This document is an application of a Pairs Trading Strategy using the Kalman Filter in order to dinamically update the hedge ratios of a pair of assets. This example is based on *Kris Longmore's* post [*"Kalman Filter Example: Pairs Trading in R*"](https://robotwealth.com/kalman-filter-pairs-trading-r/) in [RobotWealth](https://robotwealth.com/).

In this sense, we'll proceed by the following steps:

1. Select a bundle of ETF brazilian traded securities;
2. Perform cointegration and unit-root testing in several pairs of such assets;
3. Perform the pairs trading analysis with each pair; and
4. Compare the results and point to which pairs are the most interesting to trade on.

```{r packages, warning=FALSE,message=FALSE}

#Loading the necessary packages

library(quantmod)
library(pracma)
library(kableExtra)
library(dplyr)
library(PairTrading)

```

## **How to find profitable pair trading opportunities?**

```{r Source,warning=FALSE,message=FALSE}

Assets<-c('BOVA11.SA',
          'BBSD11.SA',
          'BRAX11.SA',
          'ECOO11.SA',
          'SMAL11.SA',
          'BOVV11.SA',
          'DIVO11.SA',
          'FIND11.SA',
          'GOVE11.SA',
          'MATB11.SA',
          'ISUS11.SA',
          'PIBB11.SA',
          'SPXI11.SA')


getSymbols(Assets[1],from='2019-01-01',to=Sys.Date()-20)

Stock<-data.frame(BOVA11.SA=na.locf(BOVA11.SA[,6]))%>%as.xts(order.by=index(BOVA11.SA))

for (i in 2:length(Assets)){
  
  Asset<-na.locf(getSymbols(Assets[i],from='2019-01-01',to=Sys.Date()-20,auto.assign = FALSE)[,6])
  
  Stock<<-cbind(Stock,Asset)
  
  }

rm(list=ls()[-c(2,5)])


for (i in 2:length(Assets)){
  
pairs<-cbind(Stock[,1],
             Stock[,i])
  
names(pairs)<-c('BOVA11',paste0(Assets[i]))
assign(paste0('G',i),plot(pairs, legend.loc=1,main=''))

}


#Check for stationarity

Tests<-data.frame(Asset=NA,PP=NA,ADF=NA,Decision=NA)

for (i in 2:length(Assets)){
  
  PP<-PP.test(as.numeric(Stock[,1]-Stock[,i]))$p.value
  ADF<-adf.test(as.numeric(Stock[,1]-Stock[,i]),k=1)$p.value

  Tests<-rbind(Tests,c(Assets[i],round(PP,4),round(ADF,4)))
}

Tests[-1,]%>%arrange(ADF)%>%mutate(Decision=ifelse(ADF<0.10,'Stationary','Non-Stationary'))%>%
  kable(row.names=FALSE)%>%kable_styling(bootstrap_options = 'responsive')


```


## **Allowing for time-varying relationship: the use of the Kalman Filter**

```{r Kalman Filter}

Analysis<-function(z){
  
plot(get(paste0('G',z)))
  
### Kalman Filter
  
x <- Stock[, 1]
y <- Stock[, z]

x$int <- rep(1, nrow(x))

delta <- 0.0001
Vw <- delta/(1-delta)*diag(2)
Ve <- 0.001
R <- matrix(rep(0, 4), nrow=2)
P <- matrix(rep(0, 4), nrow=2)

beta <- matrix(rep(0, nrow(y)*2), ncol=2)
y_est <- rep(0, nrow(y))
e <- rep(0, nrow(y))
Q <- rep(0, nrow(y))

for(i in 1:nrow(y)) {
  if(i > 1) {
    beta[i, ] <- beta[i-1, ] # state transition
    R <- P + Vw # state cov prediction
  }
  y_est[i] <- x[i, ] %*% beta[i, ] # measurement prediction
  Q[i] <- x[i, ] %*% R %*% t(x[i, ]) + Ve # measurement variance prediction
  
  # error between observation of y and prediction
  e[i] <- y[i] - y_est[i]
  K <- R %*% t(x[i, ]) / Q[i] # Kalman gain
  
  # state update
  beta[i, ] <- beta[i, ] + K * e[i]
  P = R - K %*% x[i, ] %*% R
}

beta <- xts(beta, order.by=index(Stock[,1]))

par(mfrow=c(2,1))

print(plot(beta[2:nrow(beta), 1], type='l', main = 'Kalman updated hedge ratio'))
print(plot(beta[2:nrow(beta), 2], type='l', main = 'Kalman updated intercept'))


# plot trade signals
e <- xts(e, order.by=index(pairs))
sqrtQ <- xts(sqrt(Q), order.by=index(pairs))
signals <- merge(e, sqrtQ, -sqrtQ)
colnames(signals) <- c("e", "sqrtQ", "negsqrtQ")
#plot(signals[3:length(index(signals))], ylab='e', main = 'Trade signals at one-standard deviation', col=c('blue', 'black', 'black'), lwd=c(1,2,2))


# more aggressive trade signals
signals <- merge(e, .5*sqrtQ, -.5*sqrtQ)
colnames(signals) <- c("e", "sqrtQ", "negsqrtQ")
print(plot(signals[3:length(index(signals))], ylab='e', main = 'Trade signals at one-standard deviation', col=c('blue', 'black', 'black'), lwd=c(1,2,2)))



# vectorised backtest 
sig <- ifelse((signals[1:length(index(signals))]$e > signals[1:length(index(signals))]$sqrtQ) & (lag.xts(signals$e, 1) < lag.xts(signals$sqrtQ, 1)), -1, 
              ifelse((signals[1:length(index(signals))]$e < signals[1:length(index(signals))]$negsqrtQ) & (lag.xts(signals$e, 1) > lag.xts(signals$negsqrtQ, 1)), 1, 0))

colnames(sig) <- "sig"

## trick for getting only the first signals
sig[sig == 0] <- NA
sig <- na.locf(sig)
sig <- diff(sig)/2
plot(sig)

## simulate positions and pnl
sim <- merge(lag.xts(sig,1), beta[, 1], x[, 1], y)
colnames(sim) <- c("sig", "hedge", 'Asset 1', 'Asset 2')
sim$posX <- sim$sig * -1000 * sim$hedge
sim$posY <- sim$sig * 1000   
sim$posX[sim$posX == 0] <- NA
sim$posX <- na.locf(sim$posX)
sim$posY[sim$posY == 0] <- NA
sim$posY <- na.locf(sim$posY)

pnlX <- sim$posX * diff(sim[, 3])
pnlY <- sim$posY * diff(sim[, 4])
pnl <- pnlX + pnlY
print(plot(cumsum(na.omit(pnl)), main="Cumulative PnL, $"))

}
```


## **Case 1: `r Assets[2]`**
```{r Case1,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=2)

```


## **Case 2: `r Assets[3]`**
```{r Case2,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=3)

```


## **Case 3: `r Assets[4]`**
```{r Case3,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=4)

```


## **Case 4: `r Assets[5]`**
```{r Case4,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=5)

```


## **Case 5: `r Assets[6]`**
```{r Case5,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=6)

```


## **Case 6: `r Assets[7]`**
```{r Case6,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=7)

```


## **Case 7: `r Assets[8]`**
```{r Case7,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=8)

```


## **Case 8: `r Assets[9]`**
```{r Case8,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=9)

```


## **Case 9: `r Assets[10]`**
```{r Case9,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=10)

```


## **Case 10: `r Assets[11]`**
```{r Case10,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=11)

```


## **Case 11: `r Assets[12]`**
```{r Case11,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=12)

```


## **Case 12: `r Assets[13]`**
```{r Case12,warnings=FALSE,message=FALSE}

par(mfrow=c(1,1))
Analysis(z=13)

```